Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-05-31 main.tex
Adjustment of figure location and size.

\begin{figure}
  \centering
  \subcaptionbox{}{
    \includegraphics[width=3in]{figures/uncontrolled-eig-vs-speeds.png}
  }
  \subcaptionbox{}{
    \includegraphics[width=3in]{figures/balance-assist-controllers-eig-vs-speeds.png}
  }
  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
    blue lines indicate the two speeds we perform experiments at. Green shaded
    region is the linear stable range.}
  \label{fig:eigenvalues}
\end{figure}

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
%
\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{
    A participant on the bicycle in the safety harness with the Bump'Em motor
    lines attached to the ends of the handlebars.
    \label{fig:participant-in-set-up}
  }{
    \includegraphics[width=70mm]{figures/participant-in-set-up.png}
  }
  \subcaptionbox{
    Top view diagram of bicycle handlebar perturbation system.
    \label{fig:setup-diagram}
  }{
    \includegraphics[width=70mm]{figures/setup-diagram.png
  }
}
\end{figure}

\begin{figure}
  \includegraphics[width=150mm]{figures/perturbation_0.png}
  \caption{Motor force and resulting  steer torque perturbation based on an 80N
  counterclockwise applied force.}
\end{figure}

@@

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
location of their front wheel on the center line of the treadmill. We define a
``fall'' on the treadmill by two criteria: the rider removes their foot from
the pedal and places it on the ground or the bicycle wheel exceeds the width of
the treadmill. Figure~\ref{fig:fall} shows an example fall. We then applied
perturbations in random directions (clockwise or counter clockwise) increasing
the magnitude by X~\si{\newton\meter} each time.  We log the magnitude that
causes the first fall to characterize that subject's nominal fall threshold.
Following this we apply 20 perturbations of random magnitudes to the cyclist
while they ride at a constant speed and record which perturbations cause a
fall. We let the cyclist rest and then perform another 20 perturbations. We
randomize whether the balance assist system is on or off during the first or
second set of perturbations.
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Thesis Fig 5.1 on pg 37}
  \label{fig:fall}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying
perturbations above and below the initial perturbation threshold, while
allowing small progression of the perturbation threshold to accommodate
learning effects. The goal of this adaptive staircase procedure is to have
participants fall for approximately 50\% of the time. The procedure
consists of twenty perturbations per condition. Participants undergo two
conditions: twenty perturbations with the balance-assist turned on and twenty
perturbations with the balance-assist turned off (in randomized order).

@@

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
%
\begin{table}
  \caption{Raw measurements}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Perturbation Order Number & \(j\) & integer & NA\\
    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

\begin{figure}
  \centering
  \subcaptionbox{}{
    \includegraphics[width=70mm]{figures/uncontrolled-eig-vs-speeds.png}
  }
  \subcaptionbox{}{
    \includegraphics[width=70mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
  }
  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
    blue lines indicate the two speeds we perform experiments at. Green shaded
    region is the linear stable range.}
  \label{fig:eigenvalues}
\end{figure}

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
at the handlebar via four inline load cells, each with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
to that described in the thesis include simplifying the controller with a
inexpensive microcontroller and the use of a simpler safety harness.
Figure~\ref{fig:perturbation_0} shows an example perturbation.
%
\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{
    A participant on the bicycle in the safety harness with the Bump'Em motor
    lines attached to the ends of the handlebars.
    \label{fig:participant-in-set-up}
  }{
    \includegraphics[width=75mm]{figures/participant-in-set-up.png}
  }
  \subcaptionbox{
    Top view diagram of bicycle handlebar perturbation system.
    \label{fig:setup-diagram}
  }{
    \includegraphics[width=65mm]{figures/setup-diagram.png
  }
}
\end{figure}

@@

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
location of their front wheel on the center line of the treadmill. We define a
``fall'' on the treadmill by two criteria: the rider removes their foot from
the pedal and places it on the ground or the bicycle wheel exceeds the width of
the treadmill. We then applied perturbations in random directions (clockwise or
counter clockwise) increasing the magnitude by X~\si{\newton\meter} each time.
Figure~\ref{fig:perturbation-sequence} shows an example reult of a
perturbation.  We log the magnitude that causes the first fall to characterize
that subject's nominal fall threshold.  Following this we apply 20
perturbations of random magnitudes to the cyclist while they ride at a constant
speed and record which perturbations cause a fall. We let the cyclist rest and
then perform another 20 perturbations. We randomize whether the balance assist
system is on or off during the first or second set of perturbations.
%
\begin{figure}
  \centering
  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. The
goal of this adaptive staircase procedure is to have participants fall for
approximately 50\% of the time. The procedure consists of twenty perturbations
per condition. Participants undergo two conditions: twenty perturbations with
the balance-assist turned on and twenty perturbations with the balance-assist
turned off.

@@

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
Figure~\ref{fig:perturbation} shows an example perturbation force measurement
compared to our Bump'Em controller command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/perturbation_0.png}
  \caption{Motor force and resulting  steer torque perturbation based on an 80N
  counterclockwise applied force.}
  \label{fig:perturbation_0}
\end{figure}

%
\begin{table}
  \caption{Raw measurements}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Perturbation Order Number & \(j\) & integer & NA\\
    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
    \bottomrule
  \end{tabular}
\end{table}
